---
title: "Analysis of Australian Athletes' BMI values"
author: "Ken Butler"
date: "October 26, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup (packages)

```{r}
library(tidyverse)
library(smmr)
```


## The Australian athletes data

Data were collected on several variables for 202 Australian elite athletes:

```{r}
athletes=read_tsv("ais.txt")
athletes
```

We will study the variable `BMI` here.

## A 90% CI for mean BMI

```{r}
t.test(athletes$BMI,conf.level=0.90)
```

The interval is from 22.62 to 23.29.

## Assessing assumptions

Look at a plot of the BMI values: are they approx normal? Draw a histogram, but first find number of bins from Sturges' rule:

```{r}
mybins=nclass.Sturges(athletes$BMI)
mybins
ggplot(athletes,aes(x=BMI))+geom_histogram(bins=mybins)
```

or a normal quantile plot:

```{r}
qqnorm(athletes$BMI)
qqline(athletes$BMI)
```

This is not approximately normal; the right tail is too long.

## CI for median from sign test

```{r}
v=ci_median(athletes,BMI,conf.level=0.90)
v
```

This goes from `r round(v[1],2)` to `r round(v[2],2)`.


